Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-04-22 paper.tex
Comment
Changes bibliography and spell check

\abstract{The aim of the experiments described in this paper is to
evaluate the use of keyboard and mouse dynamics as an appropriate
non-obtrusive sensory input for an system that is sensitive to the
affective state of its user. Our motivation for starting this research
line has been the lack of tools and methodologies for taking into
account this affective state in learning
environments. In an ideal situation, when instructors have to choose
from  a collection of programming assignments, they should consider the
  student\'s affective state and skills in order to select a learning resource
  with the appropriate level of difficulty. However, neither the data
  or the ability to process it are present in current learning
  management systems. This work tries to address this problem, by
  focused on the capture and pre-processing of data that is going to
  be fed to several machine learning techniques with the objective of classifying the affective
states of users with different levels of expertise
when learning a new programming language. We capture student data from
a web-based platform, a learning management system where students
interact whith programming excersises. We introduce in this paper a
series of pre-processing techniques that are able to convert data from keyboard and mouse dynamics captured from students as
they were coding basic Python programming assignments  into feature
vector, that have been later used for the classification into five affective states:
boredom, frustration, distraction, relaxation and engagement.
The following classification algorithms have been evaluated: k-nearest
neighbors,feed forward neural networks, na\"ive Bayes classifier, J-48
tree induction algorithm, deep learning, random forest, gradient boosted trees and na\"ive Bayes Kernel). The best accuracy was around 78\% and was achieved
by the tree induction algorithms. Results show that data gathered from ready-available,
non-obtrusive sensors can be used successfully
as another input to hyebrid classification models in order to predict an
individual\'s affective states, and these in turn can be used to make
students more engaged and thus learn more efficiently.}

@@

In order to promote engagement, sometimes instruction material is designed with the
objective of creating learning activities that are challenging to students, but
not so much as to frustrate them. Engagement is not only achieved through the
design of single learning activities but also on the appropiate selection and
sequence in which they are presented.
For this, experienced instructors gauge a student\'s affective state and then
assign an activity with the appropriate level of difficulty. However,
recognition of emotions is a difficult task, even
for humans. Instructors often use their social intelligence to recognize
students\'' affective states. During presential  classes, an instructor habitually reads the
faces of students in the classroom to see if they are confused, bored or
engaged and then decides what to do next. However, that is impossible
in asynchronous and virtual learning environments, that is why there
have been several ways of including automatic perception of emotions
in Interactive Learning Environments (ILEs) and Learning Management
Systems \cite{gottardo2017affective} and even in Integrated
Development Environments \cite{hundhausen2017ide}.

But this of course presupposes that the affective state of the student
is available to the system. This detection and recognition can be done
in real time in different ways. In the wider context of the detection
of emotions and affective 
state, many researchers have
demonstrated that affect-aware computers can provide a better performance in
assisting humans
\cite{picard2001toward,munson2018models,bosch2017affective}, and that
providing a richer experience while learning to program might improve
the outcome by making the students more engaged with it
\cite{asensio2014progamer}. However, when ILEs embrace these methodologies
and techniques a
problem arises: Sensors which are not present in the default computer
configuration must be used to gather data that will be used to perform the
recognition of users\' affective states.
Some sensors rely on physiological readings
and must be in physical contact with students \cite{atkinson2017assessing,swansianalyzing}.
These sensors can be considered
intrusive or invasive, not to mention that you have to acquire them
and convice the student to wear them while learning.  They can also disrupt the student\'s learning experience
\cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
Other less cumbersome sensors, such as video cameras, which are in
fact default equipment in most laptops, can also be considered invasive, since
they always transmit the user's identity, appearance and behavior along with
the emotional data \cite{picard2001toward}. So what we would need in
order to embed recognition of affective states in learning systems
would be non-invasive, anonymous and also ubiquitous sensors. In this
paper we are going to test such a kind of system.

And in order to do that, we will use the dynamic of the learning
environment itself. In programming courses, students solve assignments
by implementing their solution in a particular language; in order to
do this, they must type the code, run it directly or by compiling it
to check its results. This cycle can be repeated many times, and as
they are in this cycle, studens often change their affective state
according to their performence.  Data can, then, be non-intrusively
collected while students are working on their assignments and typing
on the keyboard and interacting with the GUI via mouse or other
pointing device.  All the dynamics of actually typing the code and ILE
feedback received by students, the corrections they tried, the elapsed
time between each step, the number of attempts, compiling errors,
exceptions, among others, are collected by the ILE and can be
used. Data mining of all this data could bring us a better
understanding of the learning process, and also a better model of the
student state for both the selection of the appropriate content to be
shown next and to evaluate the learner\'s competencies. But the
problem that remains is to how to recognize the student's affective
state so that it can be used by an adaptive system.

In this work, a method for the recognition of affective states in real
time through the
mining of keystroke and mouse dynamics data is analyzed further after
its initial presentation in \cite{ijcci17}.
Since students can vary the dynamic of keystrokes according
to their affective state when programming, a correlation between the
user\'s interactions (keyboard and mouse) with an emotional state has been
established in preliminary work by Zimmermann
et al. \cite{zimmermann2003affective}, but the work experiments where
conducted by asking users to write a predefined message, not programming
exercises.
In order to test the correlation of keyboard and mouse dynamics and an
affective state when programming, an interactive web-based
programming platform was initially proposed by us in
\cite{ijcci17}. In this paper we complete the background and
motivation for this research, and update the state of the art, as well
as justify better the methdology used and the way data has been
collected, extending the results and the methods used. The platform we
presented in that paper originally was 
designed to collect keyboard and mouse data 
from the students\' interaction for data analysis. The proposed method
has three main components: a JavaScript library to capture keyboard
and mouse dynamics from a browser-based editor, a pre-processing step
whose output is a feature vector that is sent to a third component,
the classification algorithm.

An experiment was conducted where students solved a series of programming
assignments using the web based learning environment. Using only the
data from the student\'s keyboard and mouse dynamics six affective
states where recognized for each of the attempts at solving the
assignments. Results obtained from the experiment are promising. The
affective states recognized include the most common in novice
programmers according to the study of Bosch, D'Mello \& Mills
\cite{bixler2013detecting}: boredom, engagement/flow, confusion and
frustration. For the initial experiment \cite{ijcci17} 
four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
classifier, feed forward neural network, and a na\"ive Bayes. The output of each
classifier determined if the student experienced the affective state during the
assignment. In this experiment we will also use 
Deep Learning \cite{candel2016deep} and  Gradient Boosted Trees 
via H2O version 3.8.2.6, and the Random Forest and Na\"ive Bayes Kernel 
implementatios from RapidMiner Studio version 8.1.

@@

Table \ref{tab_performance} shows the performance of each of the four
classifiers together
with the $\kappa$ coefficients. A 10-Fold cross validation was used.
The accuracies and $\kappa$ coefficients obtained are close to what is
usually obtained in fixed-text methods, for example, the results presented in
Epp, Lippold y Mandryk \cite{epp2011identifying}.
In a fixed-text method, subjects are asked to write specific texts.
The results obtained with this method are satisfactory, considering that
learners were writing a program, a task comparable with free-text writing,
where there is no restriction on what has to be written.
In the case of the $\kappa$ coefficients, it is
usual to see values below 0.2 in methods involving free-text. In this case, some
values of $\kappa$ were close to 0.5, is important to consider the values of $\kappa$
in these results because the class distribution is not uniformly distributed. As
it is observed in many works in affective recognition, decision trees normally
produce competitive accuracy. In this case the J-48 classifier obtained an
accuracy of 80.48\%, which is marginally better than the others, and an accepted $\kappa$
statistic for the kind of problem. The artificial neural network (ANN)
gives the highest
accuracy in the classification of frustration.
%
\begin{table}[h!tb]
\centering
\caption{Parameters used in the evolutionary algorithm for feature
  weight optimization.}
\label{tab_ga_w}
    \begin{tabular}{ | l | l | }
      \hline
      Parameter & Value \\
      \hline
      Initialize probability & 0.5 \\
      Population size & 10 \\
      Generations & 30 \\
      Selection & Tournament \\
      Tournament size & population size / 4 \\
      Crossover & Uniform \\
      Crossover probability & 0.5 \\
      Mutation rate & $1 \over 39$ \\
      Minimum number of features & 5 \\
        \hline
    \end{tabular}
  \end{table}
% 
\begin{table*}[!t]
\centering
\caption{Parameters used for the different classification algorithms
  used for recognizing the Flow/engaged state.}
\label{tab_params_ml}
    \begin{tabular}{ | l | p{10cm} |  }
\hline
      Algorithm & Parameters \\
      \hline 
      Random Forest &  number of trees = 10, 
criterion = gain ratio, maximal depth = 20, 
                      confidence = 0.25, minimal gain = 0.1, minmail leaf size = 2 \\
      J48 &  confidence = 0.79, minimal leaf size=2 \\
      kNN & $k$ = 4 \\
      Gradient boosted trees & number of trees = 20,  maximal depth = 5, minimum rows = 10, 
minimum split 
improvement = 0.0, number of bins = 20, 
                               learning rate = 0.1, sample rate = 1.0 \\
      Bayes Kernel & laplace correction = True, estimation mode = greedy, 
                     minimum bandwidth = 0.1, number of kernels = 10 \\
      Deep learning & activation = Rectifier, hidden layer sizes =
                      [50, 50], epochs = 10 \\
      \hline
    \end{tabular}
  \end{table*}

Table \ref{tab_perf_flow} shows the performance for the additional classifiers
evaluated specifically for the Flow/engaged affective state. The goal was to
increase the performance achieved earlier. Again a ten-
fold cross-validation was conducted. Again, an evolutionary algorithm was used for feature with the parameters shown in Table \ref{tab_ga_w}.
      % The
% parameters used for the evolutionary optimization algorithm are the following:
% Population size = 10, generations = 30, selection = tournament, tournament size = 0.25 (fraction of the population),  initialize probability = 0.5,  mutation
%       probability = 1/number of attributes,  crossover probability =
%       0.5.
      % Please Mario check these parameters and how they have put to
      % the table; I don't know what's the "initialization
      % probability" - JJ
     % Is the probability a feature has to be "on" (used by the algorithm)
     % initially.
% Does that mean that you make probablity = 0.5? Still don't
% understand, but I have added to the table - JJ
      The parameters used in this case for the classification
      algorithms, and the algorithms used are shown in Table \ref{tab_params_ml}
% used for the Random Forest algorithm: number of trees = 10, 
% criterion = gain ratio, maximal depth = 20, 
% confidence = 0.25, minimal gain = 0.1, minmail leaf size = 2.  The
% parameters used for are the J48 are the following:  confidence = 0.79, minimal
% leaf size=2. For the kNN algorithm, the value k = 4 was used. 
% For gradient boosted
% trees: number of trees = 20,  maximal depth = 5, minimum rows = 10, 
% minimum split 
% improvement = 0.0, number of bins = 20, 
% learning rate = 0.1, sample rate = 1.0.  Bayes
% Kernel: laplace correction = True, estimation mode = greedy, 
% minimum bandwidth = 0.1, number of kernels = 10.  Finally for deep learning: activation = Rectifier, hidden layer sizes = [50, 50], epochs = 10.

@@

The aim of this research was to evaluate the use of keyboard and mouse dynamics
as an appropriate sensory input for an affective recognition system. The context
of use is a learning environment for programmers, in particular for learning
assignments consisting in writing short programs interactively. An experimental
approach was adopted with this objective, by effectively using the
method with real subjects using the environment.

The proposed method in this work obtained
satisfactory results. It is usual for a classification method based on free-text
to obtain an accuracy and $\kappa$ measure below their counterparts of
fixed-text; however, in this
work the results obtained were similar to those works applied to fixed-text
dynamics. A possible explanation of this would be the addition of the mouse
dynamics features and the additional preprocessing performed on the
feature vectors.  Out of all methods tested, tree induction algorithms
obtained better results when we take into consideration both accuracy and  $\kappa$ values.

While these are promising results further experiments are
needed. Other experiments should focus on novice programmers, with
which we expect to gather more data and more representative examples for all the
clasess. These experiments should give researchers the hability to do stratified samples to balence
the class distribution and achieve better results. Also more experiments should
give more insigth on the relationship between affective state  and interactive
behavior. Results show near 80\% accuracy, these results  are competitive
against other classifiers of a similar domain, but we do not know yet what is
the effectiveness when applied to affective-aware learing environments.

%This
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text?
% Is 80% effective?
%  - done  Mario 
Future work will focus on extracting other features from the learners\' interaction such as behavioral data, as proposed by Blikstein
\cite{blikstein2011using}; the parameters proposed by this author
include the size of the program and changes in the code. Other authors
also propose treating the use of certain keys as features e.g., the
delete key because it is mainly used for correcting sentences. Also a multi-modal approach is needed. Since all these need a completely new approach, a new methodology will be developed to
gather data and take measures.

@@

\bibliographystyle{apalike}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{programming}

>>>>

\abstract{The aim of the experiments described in this paper is to
evaluate the use of keyboard and mouse dynamics as an appropriate
non-obtrusive sensory input for an system that is sensitive to the
affective state of its user. Our motivation for starting this research
line has been the lack of tools and methodologies for taking into
account this affective state in learning
environments. In an ideal situation, when instructors have to choose
from  a collection of programming assignments, they should consider the
  student\'s affective state and skills in order to select a learning resource
  with the appropriate level of difficulty. However, neither the data
  or the ability to process it are present in current learning
  management systems. This work tries to address this problem, by
  focused on the capture and pre-processing of data that is going to
  be fed to several machine learning techniques with the objective of classifying the affective
states of users with different levels of expertise
when learning a new programming language. We capture student data from
a web-based platform, a learning management system where students
interact with programming exorcises. We introduce in this paper a
series of pre-processing techniques that are able to convert data from keyboard and mouse dynamics captured from students as
they were coding basic Python programming assignments  into feature
vector, that have been later used for the classification into five affective states:
boredom, frustration, distraction, relaxation and engagement.
The following classification algorithms have been evaluated: k-nearest
neighbors,feed forward neural networks, na\"ive Bayes classifier, J-48
tree induction algorithm, deep learning, random forest, gradient boosted trees and na\"ive Bayes Kernel). The best accuracy was around 78\% and was achieved
by the tree induction algorithms. Results show that data gathered from ready-available,
non-obtrusive sensors can be used successfully
as another input to hybrid classification models in order to predict an
individual\'s affective states, and these in turn can be used to make
students more engaged and thus learn more efficiently.}

@@

In order to promote engagement, sometimes instruction material is designed with the
objective of creating learning activities that are challenging to students, but
not so much as to frustrate them. Engagement is not only achieved through the
design of single learning activities but also on the appropriate selection and
sequence in which they are presented.
For this, experienced instructors gauge a student\'s affective state and then
assign an activity with the appropriate level of difficulty. However,
recognition of emotions is a difficult task, even
for humans. Instructors often use their social intelligence to recognize
students\'' affective states. During presential  classes, an instructor habitually reads the
faces of students in the classroom to see if they are confused, bored or
engaged and then decides what to do next. However, that is impossible
in asynchronous and virtual learning environments, that is why there
have been several ways of including automatic perception of emotions
in Interactive Learning Environments (ILEs) and Learning Management
Systems \cite{gottardo2017affective} and even in Integrated
Development Environments \cite{hundhausen2017ide}.

But this of course presupposes that the affective state of the student
is available to the system. This detection and recognition can be done
in real time in different ways. In the wider context of the detection
of emotions and affective 
state, many researchers have
demonstrated that affect-aware computers can provide a better performance in
assisting humans
\cite{picard2001toward,munson2018models,bosch2017affective}, and that
providing a richer experience while learning to program might improve
the outcome by making the students more engaged with it
\cite{asensio2014progamer}. However, when ILEs embrace these methodologies
and techniques a
problem arises: Sensors which are not present in the default computer
configuration must be used to gather data that will be used to perform the
recognition of users\' affective states.
Some sensors rely on physiological readings
and must be in physical contact with students \cite{atkinson2017assessing,swansianalyzing}.
These sensors can be considered
intrusive or invasive, not to mention that you have to acquire them
and convince the student to wear them while learning.  They can also disrupt the student\'s learning experience
\cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
Other less cumbersome sensors, such as video cameras, which are in
fact default equipment in most laptops, can also be considered invasive, since
they always transmit the user's identity, appearance and behavior along with
the emotional data \cite{picard2001toward}. So what we would need in
order to embed recognition of affective states in learning systems
would be non-invasive, anonymous and also ubiquitous sensors. In this
paper we are going to test such a kind of system.

And in order to do that, we will use the dynamic of the learning
environment itself. In programming courses, students solve assignments
by implementing their solution in a particular language; in order to
do this, they must type the code, run it directly or by compiling it
to check its results. This cycle can be repeated many times, and as
they are in this cycle, students often change their affective state
according to their performance.  Data can, then, be non-intrusively
collected while students are working on their assignments and typing
on the keyboard and interacting with the GUI via mouse or other
pointing device.  All the dynamics of actually typing the code and ILE
feedback received by students, the corrections they tried, the elapsed
time between each step, the number of attempts, compiling errors,
exceptions, among others, are collected by the ILE and can be
used. Data mining of all this data could bring us a better
understanding of the learning process, and also a better model of the
student state for both the selection of the appropriate content to be
shown next and to evaluate the learner\'s competencies. But the
problem that remains is to how to recognize the student's affective
state so that it can be used by an adaptive system.

In this work, a method for the recognition of affective states in real
time through the
mining of keystroke and mouse dynamics data is analyzed further after
its initial presentation in \cite{ijcci17}.
Since students can vary the dynamic of keystrokes according
to their affective state when programming, a correlation between the
user\'s interactions (keyboard and mouse) with an emotional state has been
established in preliminary work by Zimmermann
et al. \cite{zimmermann2003affective}, but the work experiments where
conducted by asking users to write a predefined message, not programming
exercises.
In order to test the correlation of keyboard and mouse dynamics and an
affective state when programming, an interactive web-based
programming platform was initially proposed by us in
\cite{ijcci17}. In this paper we complete the background and
motivation for this research, and update the state of the art, as well
as justify better the methodology used and the way data has been
collected, extending the results and the methods used. The platform we
presented in that paper originally was 
designed to collect keyboard and mouse data 
from the students\' interaction for data analysis. The proposed method
has three main components: a JavaScript library to capture keyboard
and mouse dynamics from a browser-based editor, a pre-processing step
whose output is a feature vector that is sent to a third component,
the classification algorithm.

An experiment was conducted where students solved a series of programming
assignments using the web based learning environment. Using only the
data from the student\'s keyboard and mouse dynamics six affective
states where recognized for each of the attempts at solving the
assignments. Results obtained from the experiment are promising. The
affective states recognized include the most common in novice
programmers according to the study of Bosch, D'Mello \& Mills
\cite{bixler2013detecting}: boredom, engagement/flow, confusion and
frustration. For the initial experiment \cite{ijcci17} 
four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
classifier, feed forward neural network, and a na\"ive Bayes. The output of each
classifier determined if the student experienced the affective state during the
assignment. In this experiment we will also use 
Deep Learning \cite{candel2016deep} and  Gradient Boosted Trees 
via H2O version 3.8.2.6, and the Random Forest and Na\"ive Bayes Kernel 
implementations from RapidMiner Studio version 8.1.

@@

Table \ref{tab_performance} shows the performance of each of the four
classifiers together
with the $\kappa$ coefficients. A 10-Fold cross validation was used.
The accuracies and $\kappa$ coefficients obtained are close to what is
usually obtained in fixed-text methods, for example, the results presented in
Epp, Lippold y Mandryk \cite{epp2011identifying}.
In a fixed-text method, subjects are asked to write specific texts.
The results obtained with this method are satisfactory, considering that
learners were writing a program, a task comparable with free-text writing,
where there is no restriction on what has to be written.
In the case of the $\kappa$ coefficients, it is
usual to see values below 0.2 in methods involving free-text. In this case, some
values of $\kappa$ were close to 0.5, is important to consider the values of $\kappa$
in these results because the class distribution is not uniformly distributed. As
it is observed in many works in affective recognition, decision trees normally
produce competitive accuracy. In this case the J-48 classifier obtained an
accuracy of 80.48\%, which is marginally better than the others, and an accepted $\kappa$
statistic for the kind of problem. The artificial neural network (ANN)
gives the highest
accuracy in the classification of frustration.
%
\begin{table}[h!tb]
\centering
\caption{Parameters used in the evolutionary algorithm for feature
  weight optimization.}
\label{tab_ga_w}
    \begin{tabular}{ | l | l | }
      \hline
      Parameter & Value \\
      \hline
      Initialize probability & 0.5 \\
      Population size & 10 \\
      Generations & 30 \\
      Selection & Tournament \\
      Tournament size & population size / 4 \\
      Crossover & Uniform \\
      Crossover probability & 0.5 \\
      Mutation rate & $1 \over 39$ \\
      Minimum number of features & 5 \\
        \hline
    \end{tabular}
  \end{table}
% 
\begin{table*}[!t]
\centering
\caption{Parameters used for the different classification algorithms
  used for recognizing the Flow/engaged state.}
\label{tab_params_ml}
    \begin{tabular}{ | l | p{10cm} |  }
\hline
      Algorithm & Parameters \\
      \hline 
      Random Forest &  number of trees = 10, 
criterion = gain ratio, maximal depth = 20, 
                      confidence = 0.25, minimal gain = 0.1, minimal leaf size = 2 \\
      J48 &  confidence = 0.79, minimal leaf size=2 \\
      kNN & $k$ = 4 \\
      Gradient boosted trees & number of trees = 20,  maximal depth = 5, minimum rows = 10, 
minimum split 
improvement = 0.0, number of bins = 20, 
                               learning rate = 0.1, sample rate = 1.0 \\
      Bayes Kernel & Laplace correction = True, estimation mode = greedy, 
                     minimum bandwidth = 0.1, number of kernels = 10 \\
      Deep learning & activation = Rectifier, hidden layer sizes =
                      [50, 50], epochs = 10 \\
      \hline
    \end{tabular}
  \end{table*}

Table \ref{tab_perf_flow} shows the performance for the additional classifiers
evaluated specifically for the Flow/engaged affective state. The goal was to
increase the performance achieved earlier. Again a ten-
fold cross-validation was conducted. Again, an evolutionary algorithm was used for feature with the parameters shown in Table \ref{tab_ga_w}.
      % The
% parameters used for the evolutionary optimization algorithm are the following:
% Population size = 10, generations = 30, selection = tournament, tournament size = 0.25 (fraction of the population),  initialize probability = 0.5,  mutation
%       probability = 1/number of attributes,  crossover probability =
%       0.5.
      % Please Mario check these parameters and how they have put to
      % the table; I don't know what's the "initialization
      % probability" - JJ
     % Is the probability a feature has to be "on" (used by the algorithm)
     % initially.
% Does that mean that you make probablity = 0.5? Still don't
% understand, but I have added to the table - JJ
% Yes, is basically a binary random selection when creating
% the population. - Mario

      The parameters used in this case for the classification
      algorithms, and the algorithms used are shown in Table \ref{tab_params_ml}
% used for the Random Forest algorithm: number of trees = 10, 
% criterion = gain ratio, maximal depth = 20, 
% confidence = 0.25, minimal gain = 0.1, minmail leaf size = 2.  The
% parameters used for are the J48 are the following:  confidence = 0.79, minimal
% leaf size=2. For the kNN algorithm, the value k = 4 was used. 
% For gradient boosted
% trees: number of trees = 20,  maximal depth = 5, minimum rows = 10, 
% minimum split 
% improvement = 0.0, number of bins = 20, 
% learning rate = 0.1, sample rate = 1.0.  Bayes
% Kernel: laplace correction = True, estimation mode = greedy, 
% minimum bandwidth = 0.1, number of kernels = 10.  Finally for deep learning: activation = Rectifier, hidden layer sizes = [50, 50], epochs = 10.

@@

The aim of this research was to evaluate the use of keyboard and mouse
dynamics (KD) as an appropriate sensory input for an affective
recognition system for students of different levels learning to
program, specifically carrying out assignments consisting in writing
short programs interactively. An experimental approach was adopted
with this objective, by effectively using the method with real
subjects using the environment and using KD measurements and survey
responses as training data for several machine learning systems.

The proposed method in this work obtained
satisfactory results. It is usual for a classification method based on free-text
to obtain an accuracy and $\kappa$ measure below their counterparts of
fixed-text; however, in this
work the results obtained were similar to those works that worked with fixed-text
dynamics. A possible explanation of this would be the addition of the mouse
dynamics features and the additional preprocessing performed on the
feature vectors. Out of all methods tested, tree induction algorithms
obtained better results when we take into consideration both accuracy
and $\kappa$ values. These methods, besides, are able to give us some
insight on what are the KD features that effectively describe the
different affective states.

While these are promising results, further experiments are
needed. Other experiments should focus on novice programmers, with
which we expect to gather more data and more representative examples for all the
classes. These experiments should give researchers the ability to do stratified samples to balance
the class distribution and achieve better results. Also more experiments should
give more insight on the relationship between affective state  and interactive
behavior. Results show near 80\% accuracy, these results  are competitive
against other classifiers of a similar domain, but we do not know yet what is
the effectiveness when applied to affective-aware learning environments.

%This
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text?
% Is 80% effective?
%  - done  Mario 
Future work will focus on extracting other features from the learners\' interaction such as behavioral data, as proposed by Blikstein
\cite{blikstein2011using}; the parameters proposed by this author
include the size of the program and changes in the code. Other authors
also propose treating the use of certain keys as features e.g., the
delete key because it is mainly used for correcting sentences. Also a
multi-modal approach is needed, using attention detectors alongside KD
dynamics to measure at least the boredom/attention state independently
of the verbal assessment by the student. Since all these need a
completely new approach, a new methodology will be developed to 
gather data and take measures.

@@

\bibliographystyle{splncs03}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{programming}
